---
title: "Data analysis for cyclistic case study"
output: html_document
Last updated: "5th November 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(R = "https://cran.r-project.org"))
```
<br>
  
### Primary goal of this analysis:
How do annual members and casual riders use Cyclistic bikes differently?


```{r, include = FALSE}
# Loading the libraries
install.packages('tidyverse')
install.packages('sets')
library(tidyverse)
library(sets)
```
  
<br>
  

#### Loading the dataset
```{r file paths}
# FILE PATHS
DATA_FILE_PATH <- '../output/process/processed_dataset.csv'
FILE_OUT_PATH <- '../output/analyse/'

df <- read_csv(DATA_FILE_PATH)
```
- The dataset has over __5 million__ observations and __13 attributes__
  
<br>
  
#### Overview of the dataset
```{r}
  head(df)
```
<br>
  
#### Renaming the columns
```{r}
df <- rename(df, bike_type = rideable_type,
                 start_time = started_at,
                 end_time = ended_at,
                 member_type = member_casual
            )
```
<br>
  
#### Adding a few new columns
- __ride_length__: difference between end_time and start_time
- __day_of_the_week__: day of the week for start_time

```{r}
df <- mutate(df, ride_length = as.numeric(difftime(end_time, start_time)),
                 day_of_the_week = format(start_time,'%A'))
```
<br>
  
#### Changing categorical variables to factors
```{r}
df$bike_type <- factor(df$bike_type)
df$member_type <- factor(df$member_type)
df$start_station_name <- factor(df$start_station_name)
df$end_station_name <- factor(df$end_station_name)

# day of the week needs to be ordered too (useful while sorting)
df$day_of_the_week <- factor(df$day_of_the_week, 
                              levels=c("Sunday", "Monday", "Tuesday", 
                                        "Wednesday", "Thursday", "Friday",
                                       "Saturday"))
```
<br>
  
### Descriptive Analysis
  
#### Difference in bike_type used by casual riders and annual members
```{r, message=FALSE}
result <- summarize(group_by(df, member_type, bike_type), count_=n())
knitr::kable(result)
```
- __docked_bike__ seems to be used by only casual members
  
  
<br>
  
  
#### Ride length central tendencies (in seconds)
```{r}
summary(df$ride_length)
```
<br>
  
#### Compare members and casual users
```{r}
result <- summarize(group_by(df, member_type),
          mean_ride_len = mean(ride_length),
          max_ride_len = max(ride_length),
          min_ride_len = min(ride_length),
          median_ride_len = median(ride_length))

knitr::kable(result)
```
- Average ride length for __casual__ riders __is more than twice__ as compared to __annual__ members
  
  
<br>
  

#### Number of rides by the day of the week
```{r, message=FALSE}
result <- arrange(summarize(group_by(df, member_type, day_of_the_week), no_of_rides = n()), day_of_the_week)

# plot
ggplot(result, aes(x = day_of_the_week, y = no_of_rides, fill=member_type, group=member_type)) + 
  geom_col(position = "dodge")+
  geom_line(aes(y = no_of_rides), color = "black", position = position_dodge(width=1), linewidth = 0.3) +
  geom_point(position = position_dodge(width=1), show.legend=FALSE) + 
  scale_y_continuous(labels = scales::number_format(scale = 1e-3, suffix = "k")) +
  labs(title = "No. of rides each day of the week", x='Day of the week', y='No. of rides')
```
  
- casual riders use cyclistic more __closer to the weekend__
- annual member use cyclistic more __during the business days__
  
  
<br>
  

#### Average ride length by day of the week
```{r, message=FALSE}
result <- summarize(group_by(df, member_type, day_of_the_week),
                    mean_ride_length = mean(ride_length))

# plot
ggplot(result, 
       aes(x = day_of_the_week, y = mean_ride_length / 60, 
           color = member_type, group = member_type)) + 
  geom_line() + 
  geom_point(show.legend=FALSE) + 
  labs(title = "Mean ride length each day of the week", x = "Day of the week",
       y = "Mean ride length (in mins)") + 
  scale_y_continuous(
    breaks = seq(0, 40, by = 5),
    labels = seq(0, 40, by = 5)
  ) +
  expand_limits(y = 0)
  
```

- __Casual__ riders use cyclistic __atleast twice as much__ as __annual__ members everyday. 
- The average ride length gradually __decreases__ as we get into the week and __increases__ as we reach the weekend
  
  
<br>
  
#### Average ride length by month
```{r, message=FALSE}
result <- summarize(group_by(df, member_type, format(start_time, '%m')), mean_=mean(ride_length))

colnames(result) <- c('member_type', 'month', 'mean_ride_length')

# plot
ggplot(result, aes( x = month, y = mean_ride_length / 60, 
                    color = member_type, group = member_type)) +
  geom_line() + 
  geom_point(show.legend = FALSE) +
  scale_y_continuous(
    breaks = seq(0, 40, by = 5),
    labels = seq(0, 40, by = 5)
  ) + 
  scale_x_discrete(
    labels = c('Jan',"Feb",'Mar',"Apr","May","Jun","Jul","Aug","Sep","Oct")
  ) +
  labs(title='Average ride length by each month', x = 'Month', y = 'Average ride length (in mins)') + 
  expand_limits(y=0)
```
<br>
  
  
#### Number of rides by month
```{r, message=FALSE}
result <- summarize(group_by(df, member_type, format(start_time, '%m')), no_=n())

colnames(result) <- c('member_type', 'month', 'no_of_rides')

# plot
ggplot(result, aes( x = month, y = no_of_rides, 
                    color = member_type, group = member_type)) +
  geom_line() + 
  geom_point(show.legend = FALSE) +
  scale_y_continuous(
    breaks = seq(0, 500000, 50000),
    labels = scales::number_format(scale = 1e-3, suffix = "k")
  ) +
  scale_x_discrete(
    labels = c('Jan',"Feb",'Mar',"Apr","May","Jun","Jul","Aug","Sep","Oct") 
  ) +
  labs(title='No. of rides by each month', x = 'Month', y = 'No. of rides') + 
  expand_limits(y=0)
```

- __Spring__ and __Summer__ are the best seasons for both casual riders and annual members
  
  
<br>